Verification of the CVM algorithm with a new Recursive Analysis Technique

### 2025-01-24 isabelle/CVM_Subsample/document/root.tex
slightly edit abstract

\begin{abstract}
In 2022, Chakraborty et al.~\cite{chakraborty2022} published a streaming algorithm for the distinct
elements problem, that deviated considerably from the state-of-the art, both due to its simplicity
and avoidance of standard derandomization techniques, while still maintaining a close to optimal
logarithmic space complexity.

In this entry, we verify its correctness using a new verification technique we discovered during the
effort, that simplifies the analysis considerably compared to the orignal proof by Chakraborty et
al. The main idea is based on the well-known notion of probabilistic recurrence relations. However,
instead of recurrence relations with respect to the state, we investigate general functions of the
state and derive conditions on the function, which preserve the reccurence relation.

The latter allows us to derive concentration bounds which is usually not possible with 
recurrence relations.

This new technique also opened up the possible design space, and we introduce a new variant of the 
CVM algorithm, that uses less space, fewer steps, and also has an additional property in addition 
to concentration: unbiasedness. This means the expected result of the algorithm is exactly equal to
the desired result. The latter is also a new property, that classic algorithms for the distinct
elements problem do not posses.
\end{abstract}

>>>>

\begin{abstract}
In 2022, Chakraborty et al.~\cite{chakraborty2022} published a streaming
algorithm (henceforth, the CVM algorithm) for the distinct elements problem
that deviated considerably from the state-of-the art, due to its simplicity
and avoidance of standard derandomization techniques, while still maintaining
a close-to-optimal logarithmic space complexity.

In this entry, we verify the CVM algorithm's correctness using a new technique
which simplifies the formal analysis considerably compared to the original
proof by Chakraborty et al.. The main idea is based on the well-known notion of
probabilistic recurrence relations. However, instead of recurrence relations
with respect to the state, we investigate general functions of the state and
derive conditions on the function which preserve the relevant recurrence
relation. This generalization allows us to derive concentration bounds which is
usually not possible with recurrence relations.

This new technique opens up the possible algorithm design space, and we
introduce a new variant of the CVM algorithm that uses less space, fewer steps,
and also has an additional property in addition to concentration: unbiasedness.
The latter property means that the new variant returns the correct distinct
element count in expectation, which is a property that neither the original CVM
algorithm nor classic algorithms for the distinct elements problem posses.
% TODO(YK): in what sense does it use less space and fewer steps?
\end{abstract}
